Scaling rates of true polar wander in convecting planets and moons

### 2016-02-22 tpw_rate.tex
Change Fr to m

In particular, the two most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
This second nondimensional number does not have a uniformly agreed-upon name, but has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, which we adopt here.
Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
It would be straightforward to include them, and they do not affect the overall treatment of this scaling.

@@

\begin{equation}
\begin{aligned}
- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
\label{eq:hydrostatic}
\end{aligned}
\end{equation}
Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
 - & \mathrm{Ra \; Fr}\; T \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \nabla P^* \\ 
&+ \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \mathrm{Ra} \; T \; \mathbf{g}
\end{aligned}
\end{equation}

@@

\begin{table}
\centering
\caption{Nondimensional numbers with approximate Earth-like values}
\label{tab:nondim}
\begin{tabular}{@{}lcccc}
Symbol &  Number & Definition & Approximate value \\
\hline
Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
Fr & Froude & $\Omega_0^2 R/g_0$ & $10^{-3}$ \\
A & aspect ratio & $R_i/R$ & $0.54$ \\
$\Gamma$ & density deficit &$ \alpha \Delta T$ & $10^{-2}$ \\
\end{tabular}
\end{table}

@@

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 \mathrm{Fr}
\end{equation}

The time constant $T_1$ is a viscous relaxation time of the planetary mantle. It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
\begin{equation}
T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }
\end{equation}
Plugging these scalings into Equation~\eqref{eq:simple_milankovitch} we find
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma \mathrm{Fr}} \Lambda_{31} \sin{2 \theta}
\label{eq:scaled_rotation}
\end{equation}

@@

The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra})$ and $\theta(\mathrm{Ra})$.

@@

The preceding results clarify the complex relationship between the Rayleigh number, the Froude number and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through 
Equation~\eqref{eq:grow_perturbation} and its relaxation through Equation~\eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{\mathrm{Fr}} \sin{2 \theta}
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence of the prefactor in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} 
on the Rayleigh number.

We then suggest that the most important parameters are the Froude number, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
Whereas the rest of the expression has a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.

@@

The most important parameters are the Rayleigh number and the Froude number.
The Froude number acts as a damper to TPW.
The dependence on the Rayleigh number is more complicated, since it is a control on both the forcing 
of TPW and the response, which act in opposite directions.
Overall, however, we expect that more vigorously convecting planets should be less rotationally stable, and experience more TPW.
This perspective allows us to consider not only the polar wandering of Phanerozoic Earth,
but also allows us to hypothesize about polar wandering during the Archean and Proterozoic, or 
on other planetary bodies.

>>>>

In particular, the two most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
This second nondimensional number does not have a uniformly agreed-upon name: it has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, 
and in the geodesy community has commonly been termed $m$ \citep[e.g.][]{nakiboglu1982hydrostatic, chambat2010flattening}, which we adopt here.
Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
It would be straightforward to include them, and they do not affect the overall treatment of this scaling.

@@

\begin{equation}
\begin{aligned}
- \nabla P_0 =  \rho_0 \mathbf{g} -  \rho_0 \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r}
\label{eq:hydrostatic}
\end{aligned}
\end{equation}
Nondimensionalizing with the parameters in Table \ref{tab:nondim} and removing the reference state we find the nondimensional momentum equation:
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2 \mathbf{u} - \mathrm{Ra} \; T \; \mathbf{g} + \mathrm{Ra} \; m \; T \;{\mitbf{\omega} \times \mitbf{\omega} \times \mathbf{r}} = 0
\end{aligned}
\end{equation}

@@

\begin{equation}
\begin{aligned}
 - & \mathrm{Ra} \; m \; T \; \mitbf{\Omega} \times \mitbf{\Omega} \times \mathbf{r} = - \nabla P^* \\ 
&+ \; \nabla \cdot \left( \eta \epsilon (\mathbf{u}) \right) - \mathrm{Ra} \; T \; \mathbf{g}
\end{aligned}
\end{equation}

@@

\begin{table}
\centering
\caption{Nondimensional numbers with approximate Earth-like values}
\label{tab:nondim}
\begin{tabular}{@{}lcccc}
Symbol &  Name & Definition & Approximate value \\
\hline
Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
$m$ & Froude & $\Omega_0^2 R/g_0 = \Omega_0^2 R^3/GM$ & $10^{-3}$ \\
A & aspect ratio & $R_i/R$ & $0.54$ \\
$\Gamma$ & density deficit &$ \alpha \Delta T$ & $10^{-2}$ \\
\end{tabular}
\end{table}

@@

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces($m$ \citep{munk1960rotation}:
\begin{equation}
(C-A) \sim I_0 m
\end{equation}

The time constant $T_1$ is a viscous relaxation time of the planetary mantle. It is frequently represented as a weighted average of the different relaxation modes \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
\begin{equation}
T_1 \sim \frac{ \eta }{ \rho_0 g_0 R} = \frac{R^2}{\kappa} \frac{\Gamma}{\mathrm{Ra} }
\end{equation}
Plugging these scalings into Equation~\eqref{eq:simple_milankovitch} we find
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim - \frac{\mathrm{Ra}}{\Gamma m} \Lambda_{31} \sin{2 \theta}
\label{eq:scaled_rotation}
\end{equation}

@@

The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $m$, and $\Gamma$.  
We consider relatively slowly rotating bodies here ($m \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
In this regime, therefore, we neglect the dependence on $m$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra})$ and $\theta(\mathrm{Ra})$.

@@

The preceding results clarify the complex relationship between the Rayleigh number, $m$, and the rates of TPW for a convecting planet.
As mantle convection redistributes mass in the planet's interior, the rotational bulge moves around to stay aligned with the principal axes of the convective moment of inertia. 
There is a constant competition between growth of the mismatch angle $\theta$ through 
Equation~\eqref{eq:grow_perturbation} and its relaxation through Equation~\eqref{eq:simple_milankovitch}.
We may plug in the estimate for $\Lambda_{ij}$ into equation~\eqref{eq:simple_milankovitch} to find the strikingly simple expression
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{1}{m} \sin{2 \theta}
\label{eq:simplest_milankovitch}
\end{equation}
Surprisingly, $\Gamma$ and $\mathrm{Ra}$ have completely dropped from the prefactor in the scaling. 
Response timescales for relaxation of the mismatch angle go down at high Rayleigh numbers.
At the same time, however, coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{\kappa}{R^2} \; \dot{\theta} \sim -\frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence of the prefactor in Equation~\eqref{eq:simplest_milankovitch_defer_scaling} 
on the Rayleigh number.

We then suggest that the most important parameters are $m$, which acts as the brakes on the system, and the mismatch angle $\theta(\mathrm{Ra})$.
Whereas the rest of the expression has a weak dependence on $\mathrm{Ra}$, $\theta$ is expected to be strongly dependent on it.

@@

The most important parameters are the Rayleigh number and the $m$, which acts as a damper to TPW.
The dependence on the Rayleigh number is more complicated, since it is a control on both the forcing 
of TPW and the response, which act in opposite directions.
Overall, however, we expect that more vigorously convecting planets should be less rotationally stable, and experience more TPW.
This perspective allows us to consider not only the polar wandering of Phanerozoic Earth,
but also allows us to hypothesize about polar wandering during the Archean and Proterozoic, or 
on other planetary bodies.
